Mapping short reads, faithfully

### 2020-02-10 preprint.tex
Version for first submission

\documentclass[a4,center,fleqn]{NAR}

% Enter dates of publication
\copyrightyear{2020}
\pubdate{31 December 2020}
\pubyear{2020}
\jvolume{XX}
\jissue{YY}

%\articlesubtype{This is the article type (optional)}

@@

\author{%
Eduard Valera Zorita\,$^1$,
Ruggero Cortini\,$^1$,
Guillaume J. Filion\,$^{1,2,3}$
\footnote{To whom correspondence should be addressed.
Email: guillaume.filion@gmail.com}}

\address{%
$^{1}$Center for Genomic Regulation (CRG), The Barcelona Institute
of Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain;
$^{2}$University Pompeu Fabra (UPF), Barcelona, Spain;
$^{3}$present address: Department of Biological Sciences, University
of Toronto Scarborough, Toronto, ON, Canada.}
% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

\history{%
Received January 1, 2009;
Revised February 1, 2009;
Accepted March 1, 2009}

@@

\begin{abstract}
Mapping is the process of finding the original location of a DNA read in a
reference sequence, typically a genome. Short read mappers are software tools
used in most applications that involve high-throughput sequencing. As such, they
must be continuously improved to keep up with increasing needs. Modern mappers
rely on fast seeding heuristics that convert them in inexact algorithms.
For lack of a method to compute the reliability of reported mapping locations,
mappers have so far used \textit{ad hoc} approximations of variable
quality. Here we devise the elements of a strategy to map short reads
faithfully, \textit{i.e.}, to provide accurate probabilities of mapping reliability.
The key is to estimate the number of paralogs of the target reference sequence
so as to identify these reads that can be mapped with high confidence.
This approach highlights the existence of a class of reads that can be
mapped with unprecedented confidence. We exploit this strategy in a
prototype mapper that is competitive with state-of-the-art mappers BWA-MEM
and Bowtie2, with the benefit of faithfulness. The software is open-source
and availble at \url{https://github.com/gui11aume/smmfdp}.
\end{abstract}

@@

High throughput DNA sequencing is now a well-established technology with
countless applications in industry and medicine~\cite{pmid30084865}. The
Illumina short-read technology currently dominates the market of DNA
sequencing over newer technologies that promise longer reads at the
expense of higher error rates. Software and tools exploiting this technology
are therefore an important target for optimization~\cite{Wikillu}.

In most analyses, sequencing reads are mapped to a known reference, typically
a genome, using a software tool known as a mapper. Mappers are complex
algorithms that must solve approximate string-matching problems, provided that
the reads and the reference do not match perfectly due to biological divergence
andn due to measurement errors of sequencing machinery. The traditional vision
of mapping algorithms is to optimize speed and accuracy while maintaining a low
memory footprint. From this perspective, a major breakthrough was the
conception of the FM-index~\cite{ferragina2000opportunistic,
ferragina2005indexing}, a data structure based on the Burrows-Wheeler
transform~\cite{burrows1994block} and the suffix
array~\cite{manber1993suffix}. Modern short-read mappers such as
BWA-MEM~\cite{li2013aligning} and Bowtie2~\cite{pmid22388286} rely on
specialized implementations of the FM-index for DNA sequences. More
recently, new specialized algorithms such as RNA-seq mappers have further improved
the speed and the accuracy for their specific applications~\cite{pmid27043002}.

@@

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{pmid19505943}, featuring a mandatory \emph{mapping
quality} field MAPQ, defined as ``$-10 \cdot \log_{10}Pr$\{mapping
position is wrong\}''. However, there is presently no exact method to
estimate these probabilities and most mappers rely on approximations
that are accurate only for some specific genomes and sequencing technologies.
The different implementations of mapping quality estimators are usually
undocumented, variable between mappers and versions, often spreading
puzzlement and frustration in the community~\cite{ACGT, QCFail}.
Some mappers, like STAR~\cite{pmid23104886}, use the MAPQ field as a
qualitative scale.

As a consequence, mapping quality is often neglected by users and
developpers, even though it is critical in many applications. For
instance, when calling \textit{de novo} mutations, the standard procedure
is to map sequencing reads to the genome of interest and identify mismatches
as evidence for mutations. In this context, mapping errors cause artifacts that
explain nonexistent mismatches,
so the confidence in the final call is commensurate with the mapping
quality. Inaccuracies of the MAPQ estimates cause recurrent errors in
cancer genomics~\cite{dunno} where faithfulness would make a major
difference.

Mapping quality is also important in multi-genome or multi-reference
applications. For instance, in experiments on chimeras, the confidence
that a read is mapped to the correct location is in fact the confidence
that the read is assigned to the correct genome. Applied to RNA-seq,
mapping quality thus dictates mono-allelic expression calls; applied to
Hi-C it dictates the inferred contact map of the chromosome. More
generally, in this context every analysis depends on the confidence in
the locations of the reads.

Improving the mapping process requires to understand why it is sometimes
inaccurate in the first place. Mapping heuristics are usually based on a
filtration strategy called \emph{seeding}, whereby short exact matches between
the read and the reference are used to extract a set of candidate
locations. Seeding is error-prone because errors in the read may cause correct
location to be filtered out. We recently developped a formal computational
framework to compute the error rate of different seeding strategies used
for mapping short reads~\cite{Filion619155}.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.6]{MEM_19.pdf}
\end{center}
\caption{Projected performance of MEM seeds. The probability of filtering
out the target is plotted against the size of the read (notice the log
scale on the $y$ axis). Computations are carried out with \texttt{sesame}
for MEM seeds of minimum size 19, assuming an error rate of 1\% and a
divergence rate of 6\% between paralogs (\textit{i.e.} the probability
that a given nucleotide of the target is different in a given paralog is
0.06). Each line represents the probabilities for a different number $N$
of paralogs. Note that the asymptotic decay is approximately constant up
to $N = 20$ and slows down for larger $N$.}
\label{fig_MEM}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.8]{decision_tree.pdf}
\end{center}
\caption{Proposed strategy for a faithful mapper. Each read is routed
through one of three mapping pathways. First, all the MEM seeds are
extracted (top). Then the number $N$ of paralogs of the target is
estimated. If $N$ is 20 or less (left branch), all the candidates are
tested with the Needleman-Wunsch algorithm. Once the best candidate is
found, the probability that it is unique is estimated. If it is high and
the alignment has no mismatch, mapping quality is computed using
expression (\ref{eq_super}). If the best candidate is not unique or if the
alignment has a mismatch, mapping quality is computed using expression
(\ref{eq_bayes}) or (\ref{eq_x0}). If $N$ is higher than 20 (right
branch), only the largest MEM seed is used and the mapping quality is
computed using expression (\ref{eq_low}).}
\label{fig_tree}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.84]{pushing_backward.pdf}
\end{center}
\caption{Estimating $N$ with the backward search. \textbf{(a)} The target
has no or few paralogs. It takes only a few iterations until the backward
search (BWS) finds fewer than 2 hits in the genome. \textbf{(b)} The
target has many paralogs. In this case it takes more iterations before the
query has fewer than 2 hits. The number of iterations above the threshold
is used as a statistic to estimate $N$.}
\label{fig_back}
\end{figure}

@@

For a complete test, the proposed location of the read is segmented in
20-mers overlapping by 10 nucleotides, as shown in
Figure~\ref{fig_supertest}b. The sequence is considered unique if all the
extracted 20-mers have a single hit in the genome.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.75]{supertest.pdf}
\end{center}
\caption{Test for uniqueness. \textbf{(a)} Basic principle. After the
alignment stage, the sequence of the best hit is segmented in 20-mers. The
uniqueness of a given 20-mer is tested with the backward search.
\textbf{(b)} Complete test. The genomic location returned by the search
process is considered to be unique if all the 20-mers overlapping by 10
nucleotides are unique.}
\label{fig_supertest}
\end{figure}

While testing this procedure, we noticed that a class of reads were mapped
with higher confidence than ever reported before. The \emph{super reads},
as we called them, were those aligning perfectly to a unique location.

The extremely high mapping quality of super reads comes from an unexpected
property of the test depicted in Figure~\ref{fig_supertest}. To see why
they can be mapped with such confidence, consider a read aligning
perfectly to an incorrect sequence of the genome. Observe that for
this to happen, the read must contain one or more errors that are perfectly
compensated in some paralog of the target. This can only happen if $i.$
the target has a paralog, $ii.$ the read contains an error, $iii.$ this
error matches the paralog and $iv.$ the target and the paralog are
otherwise identical.

\begin{figure*}[t]
\begin{center}
\includegraphics[scale=.79]{super_reads.pdf}
\end{center}
\caption{Properties of super reads. \textbf{(a)} Incorrect perfect
alignment. For a read to align perfectly to an incorrect sequence, the
read must contain an error (star in the read) and the incorrect nucleotide
must match a paralog (star in the hit). This is not a super read because
some extracted 20-mers are not unique (red arrows). \textbf{(b)} Incorrect
perfect alignment with several errors. Adding compensated errors is not
always sufficient to make 20-mers unique. To much space between errors can
leave 20-mers that are not unique, in which case the read is not super.
\textbf{(c)} Incorrectly mapped super reads. If the compensated errors are
in odd-numbered 10-mers (grey boxes), then all the extracted 20-mers are
unique and the read is super. Those conditions are exceedingly rare, so
super reads are mapped with high confidence.}
\label{fig_supread}
\end{figure*}

But those conditions are not sufficient to make a read \emph{super}. As
shown in Figure~\ref{fig_supread}a, the mapping location would not be
considered unique because there would exist at least one 20-mer where the
target and its dupliate are identical. For a read of size 50, the
paralog can be overlooked only if there are at least two errors in the
read that match the paralog.

But as shown in Figure~\ref{fig_supread}b, this is still not sufficient
for a wrongly mapped read to qualify as \emph{super}. For this, every
second 10-mer must contain an error that is exactly compensatd in the
paralog, as indicated in Figure~\ref{fig_supread}c. In this
configuration, the errors mask the presence of a paralog on all the
tested 20-mers and the incorrect target location is considered unique.
There are other scenarios where a super read is mapped to an incorrect
location, but they involve even more compensated errors so they can be
neglected.

@@

We implemented the mapping strategy presented in Figure~\ref{fig_tree} in
a prototype mapper called MEM Mapper Prototype (MMP). MMP is implemented
as a stand-alone C program including the code of
\texttt{sesame}~\cite{Filion619155} and of
\texttt{divsufsort}~\cite{fischer2017dismantling}. We used plain
mapping algorithms to better highlight the benefits of building a mapper
for faithfulness. The index consists of a standard FM-index of the genome
and its reverse complement~\citep[the implementation is detailed in
ref.][]{blog}. We also added an auxiliary lookup table storing the states
of the bacwkard search for all possible 12-mers, allowing us to skip the
first 12 iterations for every query. This lookup table has a fixed size of
256~MB. Sequence alignment is carried out using a version of the
Needleman-Wunsch algorithm~\cite{pmid5420325} with an option to abort if
the score of the current best hit is exceeded (continuing the algorithm is
useless to find the optimal location of the read). Mapping quality is
computed as detailed in the Materials and Methods section.

The implementation otherwise follows the flow chart of
Figure~\ref{fig_tree} using the formulae above, with two differences.
The first is that super reads can have 0, 1 or two mismatches, as long as
all the extracted 20-mers have a single hit in the genome. How mapping
quality is computed in those cases is described in the Supplementary
Material. The second difference is that once the best hit is found, $N$ is
estimated again using the genomic sequence instead of the read, and the
higher of the two estimates is kept.

@@

\subsubsection{Genomes and data sets}

The main characteristic of the genomes used for benchmark
(\textit{Drosophila melanogaster}, \textit{Homo sapiens} and \textit{Pinus
taeda}) are summarized in Table~\ref{table_genomes}.

\begin{table}[b]
\tableparts{%
\caption{Features of the genomes used for benchmark.}
\label{table_genomes}%
}{%
\begin{tabular*}{\columnwidth}{@{}llrr}
\toprule
Organism & Reference & Size & Sequences \\
\colrule
\textit{Drosophila melanogaster} & dm4/R6 & 150~Mbp & 1,870 \\
\textit{Homo sapiens} & hg38/GRCh38 & 3.27~Gbp & 455 \\
\textit{Pinus taeda} & GCA\_000404065.3 & 22.50~Gbp & 1,760,464 \\
\botrule
\end{tabular*}%
}
{}
\end{table}

The data sets used for benchmark consist of 50 million sequences drawn
uniformly at random from each genome. To simulate sequencing errors,
nucleotides were randomly substituted with a fixed probability; there were
no insertions or deletions. For each data set, the sequences have the same
length (50, 75 or 100 nucleotides) and the same error rate (1\%, 2\%, 5\%
or 10\%).

\subsubsection{Software}

We used BWA-MEM version 0.7.9a-r786~\cite{li2013aligning}, Bowtie 2
version 2.3.5.1~\cite{pmid22388286}, and MMP version 1.0. The genomes were
indexed with default parameters for each mapper. For the benchmark proper,
all the mappers were run in single thread. BWA-MEM and Bowtie2 were always
used with default options, and the \texttt{-e} flag of MMP was used to
specify the error rate of the data set, \textit{e.g.}, \texttt{-e .02} for
a 2\% error rate.

\subsubsection{Hardware}

All the mapping experiments were run on a Hewlett Packard Z800 workstation
with 64~bit Intel Xeon CPU X5675 at 3.07~GHz, with 64~GB of DDR3 RAM,
12~MB of cache, and running Linux Ubuntu 18.04.3. Indexing the pine genome
with MMP had to be done on a machine with 512~GB of RAM because of the
high memory footprint used by the library \texttt{divsufsort}.

\subsubsection{Benchmark results}
We set up a test using the genomes of \textit{Drosophila melanogaster}
(the fruit fly), \textit{homo sapiens} (modern humans) and \textit{Pinus
taeda} (a north American species of pine). The high-level features of the
genomes are summarized in Table~\ref{table_genomes}. We included
\textit{Drosophila} and human as model organisms with high-quality genome
assemblies, and pine as a non-model organism with a very complex genome
and a draft assembly.

To set a baseline for comparison we used BWA-MEM and Bowtie2. The purpose
of the benchmark is to test faithfulness as a viable mapping strategy, not
to run a comprehensive survey of each mapper on the data set. Each mapper
was thus used with its default parameters (but we still passed the
non-optional error rate to MMP).

\begin{figure*}[t]
\begin{center}
\includegraphics[scale=.24]{fig_faith.pdf}
\end{center}
\caption{Faithfulness with 1\% error rate. The observed mapping quality
(MAPQ) is plotted against the value computed by MMP, BWA-MEM or Bowtie2.
The data is showed for reads of size 100, 75 or 50, sampled from the
genome of the fruit fly (\textit{D. melanogaster}), of modern humans
(\textit{H. sapiens}) or of the pine (\textit{P. taeda}). The size of the
circle is proportional to the number of reads in the given category. Pink
circles indicate that no mapping error was observed, so that the observed
mapping quality was undefined. In this case one pseudo error was added if
the expected number of errors was higher than 1 (we should observe errors
but we did not), otherwise the observed mapping quality was set to the
claimed mapping quality (we should not observe errors and we did not).}
\label{fig_faith}
\end{figure*}

We measured faithfulness by comparing observed versus claimed MAPQ score
on simulated data. Figure~\ref{fig_faith} shows a series of scatter plots
at 1\% simulated error rate. The area of the circles is proportional to
the number of reads in the given category, and pink circles indicate that
the observed mapping quality had to be imputed because all the reads were
mapped correctly.

The faithfulness of MMP compares favorably to that of BWA-MEM and Bowtie2
because on average, the points lie closer to the diagonal. Importantly,
MMP explores a wider range of mapping qualities, allowing it to map reads
with extremely high confidence. Those high-confidence reads with MAPQ
above 60 are all super reads, and Figure~\ref{fig_faith} shows that in
typical mapping conditions, their number can be very high (up to two
thirds when mapping 100-mers in the \textit{Drosophila} genome).
Intuitively, the number of super reads goes down as the complexity of the
genome increases.

Similar analyses for reads with higher error rates are shown in
Supplementary Figure~\ref{fig_supp}. The faithfulness of MMP is equally
high at 2\% error rate, but it decreases substantially at 5\% and 10\%
error rates, showing that our strategy is valid only within the scope of
current short-read sequencing technologies.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=.54]{fig_acc.pdf}
\end{center}
\caption{Mapping accuracy with 1\% error rate. Plotted in each panel is
the cumulative fraction of the reads above a certain mapping quality for
MMP, BWA-MEM or Bowtie2. The score is computed empirically by sorting the
reads on MAPQ and computing the average mapping quality of the reads in
the top $x$\% for all values of $x$. The data is showed for reads of size
100, 75 or 50, sampled from the genome of the fruit fly (\textit{D.
melanogaster}), of modern humans (\textit{H. sapiens}) or of the pine
(\textit{P. taeda}).}
\label{fig_acc}
\end{figure}

Finally, it is important to evaluate the impact of faithfulness in terms
of accuracy, speed and memory footprint. Figure~\ref{fig_acc} shows that
the accuracy of MMP is competitive with that of BWA-MEM and Bowtie2
(recall that the mappers were used with default parameters). On the
\textit{Drosophila} genome, MMP is more accurate at most confidence
levels, whereas there are more variations on the human and the pine
genomes. The worst case for MMP is that of 50-mers mapped in the pine
genome, perhaps because a MEM-only seeding strategy is not competitive in
such cases. Since the goal was never to optimize accuracy, those results
show that faithfulness is compatible with accuracy.

>>>>

\documentclass[english]{article}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage[margin=1.3in]{geometry}
\usepackage{graphicx}
\usepackage[square,numbers]{natbib}
\usepackage[hyphens]{url}
\usepackage{hyperref}

@@

\author[1]{Eduard~Valera~Zorita}
\author[1]{Ruggero~Cortini}
\author[1,2,3,*]{Guillaume~J.~Filion}

\affil[1]{Center for Genomic Regulation (CRG), The Barcelona Institute
of Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain}
\affil[2]{University Pompeu Fabra (UPF), Barcelona, Spain}
\affil[3]{present address: Department of Biological Sciences, University
of Toronto Scarborough, Toronto, ON, Canada.}
\affil[*]{Correspondence: guillaume.filion@gmail.com}

@@

\begin{abstract}
Mapping is the process of finding the original location of a DNA read in a
reference sequence, typically a genome. Short read mappers are software
tools used in most applications that involve high-throughput sequencing.
As such, they must be continuously improved to keep up with increasing
needs. Modern mappers rely on seeding heuristics, making them fast but
inexact. For lack of a method to compute the reliability of their own
output, mappers have so far used approximations of variable quality. Here
we focus on faithfulness, the capacity to provide accurate mapping
confidence, and we devise a strategy to map short reads faithfully. The
key is to estimate the repetitiveness of the target reference, which is
the dominant factor for the reliability of the mapping process. This
approach highlights the existence of a class of reads that can be mapped
with unprecedented confidence. We exploit this strategy in a prototype
mapper that is competitive with state-of-the-art mappers BWA-MEM and
Bowtie2, with the benefit of faithfulness. The software is open-source and
available for download at
\url{https://github.com/gui11aume/mmp}.
\end{abstract}

@@

High throughput DNA sequencing is now a well-established technology with
countless applications in industry and medicine~\cite{pmid30084865}. The
Illumina short-read technology currently dominates the market of DNA
sequencing over newer technologies that promise longer reads at the
expense of higher error rates. Software and tools to process short-read
sequencing data are therefore an important target for
optimization~\cite{Wikillu}.

In most analyses, short reads reads are mapped to a known reference,
typically a genome, using a software tool known as a mapper. Mappers are
complex algorithms that must solve approximate string-matching problems,
where the reads and the reference do not match perfectly due to
biological divergence or to measurement errors of the sequencing
machinery. The traditional vision of mapping algorithms is to optimize
speed and accuracy while maintaining a low memory footprint. From this
perspective, a major breakthrough was the conception of the
FM-index~\cite{ferragina2000opportunistic, ferragina2005indexing}, a data
structure based on the Burrows-Wheeler transform~\cite{burrows1994block}
and the suffix array~\cite{manber1993suffix}. Modern short-read mappers
such as BWA-MEM~\cite{li2013aligning} and Bowtie2~\cite{pmid22388286} rely
on specialized implementations of the FM-index for DNA sequences. More
recently, new specialized algorithms such as RNA-seq mappers have further
improved the speed and the accuracy for their specific
applications~\cite{pmid27043002}.

@@

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{pmid19505943}, featuring a mandatory \emph{mapping
quality} field MAPQ, defined as ``$-10 \cdot \log_{10}Pr$\{mapping
position is wrong\}''. The concept was originally introduced in the MAQ
mapper~\cite{pmid18714091}, but the proposed method cannot be carried to
more modern mappers based on the FM-index.

More generally, there is no exact method to estimate these probabilities
and most mappers rely on approximations that are valid only for some
genomes and some sequencing technologies. The computations of mapping
quality scores are usually undocumented, variable between mappers and
versions, often spreading puzzlement and frustration in the
community~\cite{ACGT, QCFail}. At times, the standard is even disregarded
as some mappers, like STAR~\cite{pmid23104886}, use the MAPQ field as a
qualitative scale instead.

As a consequence, mapping quality is often neglected by users and
developpers, even though it is critical in many applications. For
instance, when calling \textit{de novo} mutations, the standard procedure
is to map sequencing reads to the genome of interest and identify
mismatches as evidence for mutations. In this context, mapping errors
cause artifacts that explain nonexistent mismatches, so the confidence in
the final call is commensurate with the mapping quality. Inaccuracies of
the mappers cause recurrent errors in cancer genomics~\cite{pmid26647970,
pmid32025007} where faithfulness would make a major difference.

Mapping quality and faithfulness are also important in multi-genome
applications. For instance, when sequencing DNA from heterozyotes or
chimeras, the confidence that a read is mapped to the correct location is
in fact the confidence that the read is assigned to the correct genome.
Applied to RNA-seq, for instance, mapping quality thus dictates the calls
for mono-allelic expression. More generally, every analysis of this type
depends on the confidence in the locations of the reads.

Improving the mapping process requires to understand why it is sometimes
inaccurate in the first place. Mapping heuristics are usually based on a
filtration strategy called \emph{seeding}, whereby short exact matches
between the read and the reference are used to extract a set of candidate
locations. The downside of seeding is that errors in the read may cause
the correct location to be filtered out, in turn the read to be mapped
incorrectly. We recently developped a formal computational framework to
compute the error rate of different seeding strategies used for mapping
short reads~\cite{Filion619155}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.6]{MEM_19.pdf}
\end{center}
\caption{Projected performance of MEM seeds. The probability of filtering
out the target is plotted against the size of the read (notice the log
scale on the $y$ axis). Computations are carried out with \texttt{sesame}
for MEM seeds of minimum size 19, assuming an error rate of 1\% and a
divergence rate of 6\% between paralogs (\textit{i.e.} the probability
that a given nucleotide of the target is different in a given paralog is
0.06). Each line represents the probabilities for a different number $N$
of paralogs. Note that the asymptotic decay is approximately constant up
to $N = 20$ and slows down for larger $N$.}
\label{fig_MEM}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.8]{decision_tree.pdf}
\end{center}
\caption{Proposed strategy for a faithful mapper. Each read is routed
through one of three mapping pathways. First, all the MEM seeds are
extracted (top). Then the number $N$ of paralogs of the target is
estimated. If $N$ is 20 or less (left branch), all the candidates are
tested with the Needleman-Wunsch algorithm. Once the best candidate is
found, the probability that it is unique is estimated. If it is high and
the alignment has no mismatch, mapping quality is computed using
expression (\ref{eq_super}). If the best candidate is not unique or if the
alignment has a mismatch, mapping quality is computed using expression
(\ref{eq_bayes}) or (\ref{eq_x0}). If $N$ is higher than 20 (right
branch), only the largest MEM seed is used and the mapping quality is
computed using expression (\ref{eq_low}).}
\label{fig_tree}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.84]{pushing_backward.pdf}
\end{center}
\caption{Estimating $N$ with the backward search. \textbf{(a)} The target
has no or few paralogs. It takes only a few iterations until the backward
search (BWS) finds fewer than 2 hits in the genome. \textbf{(b)} The
target has many paralogs. In this case it takes more iterations before the
query has fewer than 2 hits. The number of iterations before hitting the
threshold is used as a statistic to estimate $N$.}
\label{fig_back}
\end{figure}

@@

For a complete test, the candidate location of the read is segmented in
20-mers overlapping by 10 nucleotides, as shown in
Figure~\ref{fig_supertest}b. The sequence is considered unique if all the
extracted 20-mers have a single hit in the genome.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.75]{supertest.pdf}
\end{center}
\caption{Test for uniqueness. \textbf{(a)} Basic principle. After the
alignment stage, the sequence of the best hit is segmented in 20-mers. The
uniqueness of a given 20-mer is tested with the backward search.
\textbf{(b)} Complete test. The genomic location returned by the search
process is considered to be unique if all the 20-mers overlapping by 10
nucleotides are unique.}
\label{fig_supertest}
\end{figure}

While testing this procedure, we noticed that a class of reads were mapped
with higher confidence than ever reported before. The \emph{super reads},
as we called them, were those aligning without mismatch to a unique
location.

The extremely high mapping quality of super reads comes from an unexpected
property of the test depicted in Figure~\ref{fig_supertest}. To see why
they can be mapped with such confidence, consider a read aligning without
mismatch to an incorrect sequence of the genome. Observe that for this to
happen, the read must contain one or more errors that are perfectly
compensated in some paralog of the target. This can only happen if $i.$
the target has a paralog, $ii.$ the read contains an error, $iii.$ this
error matches the paralog and $iv.$ the target and the paralog are
otherwise identical.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.63]{super_reads.pdf}
\end{center}
\caption{Properties of super reads. \textbf{(a)} Incorrect alignment with
no mismatch. For a read to align perfectly to an incorrect sequence, it
must contain an error (star in the read) and the incorrect nucleotide must
match a paralog (star in the hit). This is not a super read because some
extracted 20-mers are not unique (red arrows). \textbf{(b)} Incorrect
alignment with several errors and no mismatch. Adding compensated errors
is not always sufficient to make 20-mers unique: too much space between
errors can leave 20-mers that are not unique, in which case the read is
not super. \textbf{(c)} Incorrectly mapped super reads. If the compensated
errors are in odd-numbered 10-mers (grey boxes), then all the extracted
20-mers are unique and the read is super. Those conditions are exceedingly
rare, so super reads are mapped with high confidence.}
\label{fig_supread}
\end{figure}

But those conditions are not sufficient to make a read \emph{super}. As
shown in Figure~\ref{fig_supread}a, the mapping location would not be
considered unique because there would exist at least one 20-mer where the
target and its dupliate are identical, yielding two hits. For a read of
size 50, the paralog can be overlooked only if there are at least two
errors in the read that match the paralog.

But as shown in Figure~\ref{fig_supread}b, this is still not sufficient
for a wrongly mapped read to qualify as \emph{super}. For this, the errors
must be in every second 10-mer, as indicated in Figure~\ref{fig_supread}c.
In this configuration, the errors mask the presence of another hit on all
the tested 20-mers and the incorrect target location is considered unique.
There are other scenarios where a super read is mapped to an incorrect
location, but they involve even more compensated errors so they can be
neglected.

@@

We implemented the mapping strategy presented in Figure~\ref{fig_tree} in
a prototype mapper called MEM Mapper Prototype (MMP). MMP is implemented
as a stand-alone C program including the code of
\texttt{sesame}~\cite{Filion619155} and of
\texttt{divsufsort}~\cite{fischer2017dismantling}. We used plain
mapping algorithms to better highlight the benefits of building a mapper
for faithfulness. The index consists of a standard FM-index of the genome
and its reverse complement~\citep[the implementation is detailed in
ref.][]{blog}. In line with the GEM mapper~\cite{pmid23103880}, we also
added an auxiliary lookup table storing the states of the bacwkard search
for all possible 12-mers, allowing us to skip the first 12 iterations for
every query. This lookup table has a fixed size of 256~MB. Sequence
alignment is carried out using a version of the Needleman-Wunsch
algorithm~\cite{pmid5420325} with an option to abort if the score of the
current best hit is exceeded (continuing the algorithm is useless to find
the optimal location of the read). Mapping quality is computed as detailed
in the Materials and Methods section.

The implementation otherwise follows the flow chart of
Figure~\ref{fig_tree}, with two differences. The first is that super reads
can have 0, 1 or two mismatches, as long as all the extracted 20-mers have
a single hit in the genome (the mapping quality is modified accordingly).
The second difference is that once the best hit is found, $N$ is estimated
again using the genomic sequence instead of the read, and the higher of
the two estimates is kept.

@@

We set up a test bed using the genomes of \textit{Drosophila melanogaster}
(the fruit fly), \textit{homo sapiens} (modern humans) and \textit{Pinus
taeda} (a north American species of pine). The high-level features of the
genomes are summarized in Table~\ref{table_genomes}. We included
\textit{Drosophila} and human as model organisms with high-quality genome
assemblies, and pine as a non-model organism with a very complex genome
and a draft assembly.

\begin{table}[b]
\centering
\caption{Features of the genomes used for benchmark.}
\label{table_genomes}%
\begin{tabular}{@{}llrr}
Organism & Reference & Size & Sequences \\
\hline
\textit{Drosophila melanogaster} & dm4/R6 & 150~Mbp & 1,870 \\
\textit{Homo sapiens} & hg38/GRCh38 & 3.27~Gbp & 455 \\
\textit{Pinus taeda} & GCA\_000404065.3 & 22.50~Gbp & 1,760,464 \\
\end{tabular}
\end{table}

The data sets used for the benchmark consist of 50 million sequences drawn
uniformly at random from each genome. To simulate sequencing errors,
nucleotides were randomly substituted with a fixed probability; there were
no insertions or deletions. For each data set, the sequences have the same
length (50, 75 or 100 nucleotides) and the same error rate (1\%, 2\%, 5\%
or 10\%).

To set a baseline for comparison we used BWA-MEM and Bowtie2. The purpose
of the benchmark is to test faithfulness as a viable mapping strategy, not
to run a comprehensive survey of each mapper on the data set. Each mapper
was thus used with its default parameters (but we passed the error rate to
MMP, \textit{e.g.}, \texttt{-e .02} for a 2\% error rate). We used BWA-MEM
version 0.7.9a-r786~\cite{li2013aligning}, Bowtie2 version
2.3.5.1~\cite{pmid22388286}, and MMP version 1.0. All the mappers were run
in single thread.

The mapping experiments were run on a Hewlett Packard Z800 workstation
with 64~bit Intel Xeon CPU X5675 at 3.07~GHz, with 64~GB of DDR3 RAM,
12~MB of cache, and running Linux Ubuntu 18.04.3. Indexing the pine genome
with MMP had to be done on a machine with 512~GB of RAM because of the
high memory footprint used by the library \texttt{divsufsort}.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.2]{fig_faith.pdf}
\end{center}
\caption{Faithfulness with 1\% error rate. The observed mapping quality
(MAPQ) is plotted against the value computed by MMP, BWA-MEM or Bowtie2.
The data is showed for reads of size 100, 75 or 50, sampled from the
genome of the fruit fly (\textit{D. melanogaster}), of modern humans
(\textit{H. sapiens}) or of the pine (\textit{P. taeda}). The size of the
circle is proportional to the number of reads in the given category. Pink
circles indicate that no mapping error was observed, so that the observed
mapping quality was undefined. In this case one pseudo error was added if
the expected number of errors was higher than 1 (we should observe errors
but we did not), otherwise the observed mapping quality was set to the
claimed mapping quality (we should not observe errors and we did not).}
\label{fig_faith}
\end{figure}

We measured faithfulness by comparing observed versus claimed MAPQ score
on simulated data. Figure~\ref{fig_faith} shows a series of scatter plots
at 1\% error rate on simulated reads. The area of the circles is
proportional to the number of reads in the given category, and pink
circles indicate that the observed mapping quality had to be imputed
because all the reads were mapped correctly.

The faithfulness of MMP compares favorably to that of BWA-MEM and Bowtie2
because on average, the points lie closer to the diagonal. BWA-MEM tends
to overestimate mapping quality in the high range (except for the human
genome) whereas Bowtie2 tends to understimate it. Importantly, MMP
explores a wider range of mapping qualities, allowing it to map reads with
extremely high confidence. Those high-confidence reads with MAPQ above 60
are all super reads, and Figure~\ref{fig_faith} shows that in typical
mapping conditions, their number can be very high (up to two thirds when
mapping 100-mers in the \textit{Drosophila} genome). Intuitively, the
number of super reads goes down as the complexity of the genome increases.

Similar analyses for reads with higher error rates are shown in
Supplementary Figures~5, 6 and 7. The faithfulness of MMP is equally good
at 2\% error rate, but it decreases substantially at 5\% and 10\% error
rates, showing that our strategy is valid only within the scope of current
short-read sequencing technologies.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.6]{fig_acc.pdf}
\end{center}
\caption{Mapping accuracy with 1\% error rate. Plotted in each panel is
the cumulative fraction of the reads above a certain mapping quality for
MMP, BWA-MEM or Bowtie2. The score is computed empirically by sorting the
reads on their MAPQ score and computing the average mapping quality of the
reads in the top $x$\% for all values of $x$. The data is showed for reads
of size 100, 75 or 50, sampled from the genome of the fruit fly
(\textit{D. melanogaster}), of modern humans (\textit{H. sapiens}) or of
the pine (\textit{P. taeda}).}
\label{fig_acc}
\end{figure}

It is important to evaluate the impact of faithfulness in terms of
accuracy, speed and memory footprint. Figure~\ref{fig_acc} shows that the
accuracy of MMP is competitive with that of BWA-MEM and Bowtie2 (recall
that the mappers were used with default parameters). On the
\textit{Drosophila} genome, MMP is more accurate at most confidence
levels, whereas there are more variations on the human and the pine
genomes. The worst case for MMP is that of 50-mers mapped in the pine
genome, perhaps because a MEM-only seeding strategy is not competitive in
such cases. At 2\% error rate, MMP shows a neat advantage in accuracy in
most conditions (Supplementary Figure~2) but at 5\% and 10\% error rates,
\textit{i.e.} outside the realm of current short read technologies, the
performance of MMP is more variable (Supplementary Figures~3 and 4). Those
results show that optimizing for faithfulness can resut in good
performance for accuracy.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.7]{fig_timem.pdf}
\end{center}
\caption{Resource usage at 1\% error rate. \textbf{(a)} Speed. The running
time is plotted for MMP, BWA-MEM or Bowtie2. The data is showed for reads
of size 100, 75 or 50, sampled from the genome of the fruit fly
(\textit{D. melanogaster}), of modern humans (\textit{H. sapiens}) or of
the pine (\textit{P. taeda}). \textbf{(b)} Memory usage. The peak memory
footprint is plotted for the different mappers with reads from different
genomes (the memory footprint is the same for all read sizes).}
\label{fig_timem}
\end{figure}
